\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-04-05 reviews/response-1.tex

\item[L100] \textcolor{blue}{Not clear where the Legendre transform enters: the viscous and frictional
dissipation can be read off from Eq. 4.}

Other reviewers didn't find this digression useful so we've removed it.

\item[Eqs. 6 and 12]\textcolor{blue}{While I understand that it is convenient to manipulate the
action to reflect the algebraic manipulations to yield the analytical SIA
solution, the break in symmetry between Eq. 6 and Eq. 12 is frustrating,
given that they both are name `gravity', and that they should in some
sense be the same regardless of which strain rates are assumed to be zero.}

These modules were implemented by two different authors, the SSA / hybrid models by myself (DRS) and the SIA model by Jess Badgeley (second author) as a project for her PhD thesis and to learn more about finite element analysis.
We recognize that there's a discrepancy here and this will be corrected in future releases of the package.

@@

\item[L188] \textcolor{blue}{This is a bit of a red herring, given that impenetrability is a natural
boundary condition on the incompressibility equation. It can go right
in the action principle, no extra Lagrange multiplier (besides pressure)
needed.}

I don't think this is entirely correct.
Imposing zero velocity at the boundary is easy, but imposing no normal flow or a fixed normal flow together with friction along the opposite directions is much more challenging.
When the boundary is flat, you can set one component of the velocity to 0.
But if the finite element basis for the velocity has degrees of freedom that are located at the mesh vertices, you have to confront the fact that the unit outward normal vector isn't uniquely defined at the mesh vertices.
In the next sentence we cited a paper from the Elmer/Ice group where they devised an ad-hoc (but still very effective) scheme for defining the normal vector at mesh vertices and thus imposing this boundary condition.
You can also do it with Nitsche's method for linear problems but to my knowledge no one has figured that out for power-law fluids.
I think that the natural boundary condition you're thinking of in most weak formulations of the Stokes equations is that the average of the pressure is zero.

\item[L367] \textcolor{blue}{I'm surprised by this. For challenging geometries, I've always needed to
stabilize even when using implicit Euler. Are you sure that implicit Euler
is unconditionally stable even for non-linear advection like this?}

See Donea and Huerta, \emph{Finite element methods for flow problems}, chapter 3, sections 4-6.
This section was also written very early on but became out of date by the time we submitted the paper.
Since then, we added an implicit version of the Lax-Wendroff scheme, which has better stability properties and higher order accuracy.
The text has been amended to reflect this change.

We did have similar experiences to what you describe when preparing the MISMIP+ test case for this paper.
The melt phase produces very high ablation rates concentrated in a small region right near the grounded line.
In our initial setup, the mesh we used didn't adequately resolve this feature.
The finite element interpolation errors can then have an oscillatory pattern that, while not directly amplified by the implicit Euler timestepping scheme as such, still persists and gives a nonsense solution.
This behavior manifests even at timesteps substantially below the CFL timestep.
High-amplitude oscillatory garbage in the thickness field can then result in unrealistically large driving stresses and crash the nonlinear solver for the velocity.
The extra terms from the Lax-Wendroff method help to diffuse out these oscillatory features.
Both schemes give good results when the mesh is sufficiently refined; the Lax-Wendroff scheme just require less refinement.
So this could be more an issue of spatial resolution than stability of the timestepping scheme.
You're right that saving users from having to think about the CFL condition doesn't completely alleviate all the difficulties and we state as much in the next sentence.

### 2021-04-05 reviews/response-2.tex

\item \textcolor{blue}{Can the authors comment on the overhead of the symbolic descriptions/manipulations done by their
framework? This sounds potentially like it would be very expensive. How does the cost compare to
automatic differentiation, for example? A broader question is: is computational performance/cost a
concern for users of icepack, or is it intended to be a “sandbox” in which performance is secondary
to being able to code up something “quick-and-dirty” for initial prototyping?}

See Rathgeber et al. 2016 for more information about the architecture of Firedrake and for performance benchmarks.
Loosely speaking, the path through the toolchain goes like this.
A user creates a symbolic description of the weak form of the PDE they want to solve.
Firedrake then computes a hash of this expression and looks to see if it has encountered this problem before.
If not, it does a long, complicated, and expensive series of transformations to generate highly optimized C code that fills the relevant matrices and vectors.
This C code is then compiled into a dynamic library for later reuse.
If the user has solved this problem before, Firedrake simply looks up the dynamic library that it already generated.
(Crucially, the only thing that matters is the symbolic shape of the problem, not the actual data that goes into it -- you don't have to do codegen all over again just because you changed the boundary conditions or forcing.)
In either case, Firedrake then calls into PETSc's scalable nonlinear equation solvers (SNES) to solve the resulting system of equations.
As with any just-in-time compiled language, performance is slow the first time the code is run, and faster ever after.
Most importantly, \textbf{the performance-critical parts are all written in C}.
Firedrake has been shown to scale up to large problems on thousands of processors.

It is possible to ruin the performance of the application by accidentally hard-coding a floating point value into a symbolic expression of a PDE and then changing that value in a loop.
For example you could easily make this mistake if you were doing adaptive timestepping on an evolutionary problem discretized via the method of lines.
The remedy is to wrap this value in a \texttt{firedrake.Constant} object.
This kind of performance regression is easily caught using \texttt{htop}.
While this is more a result of programmer error, it's an easy mistake to make and the Firedrake team are working on ways to diagnose it and issue appropriate warnings.

@@

For the problems that we have used icepack for so far, we have focused more on individual glaciers or catchments, and thus performance has been a secondary concern.
We aim to move towards larger continental-scale problems in the future.
The rate-limiting factor there is more our ability to find the right incantation of PETSc solver options and preconditioners than it is any inherent limitation in our tools.

\item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a ``hybrid
model''. I see that it is hybrid in the sense that you have a different discretization in the horizontal and
vertical direction, but there are also hybrid ice models that use different PDEs in different domains,
e.g., the ISCAL model of Ahlkrona et al. (https://doi.org/10.1016/j.quascirev.2016.01.032).
Is the term ``hybrid model'' a common name for the approach in Section 2.2.3? Perhaps it is and I
am not aware of it.
Does the hybrid model described in the paper have the same applicability as say the First Order
Stokes model? Can it be used for both Greenland and Antarctica at continental scales?}

The fundamental physics are the first-order model obtained by asymptotic expansion of the Stokes equations in the aspect ratio, also known as the Blatter-Pattyn equations.
What I had imagined is that using only vertical basis functions up to degree 2 essentially defines its own semi-discrete physics model, similar to two- or three-layer ocean models.
You can view these as very coarse discretizations of the primitive equations, or you can view them as simplified models in their own right.
But I made this naming choice before there were many other collaborators on the project.
The ensuing confusion has shown that this was a bad choice of terminology and we intend to change it in a future version.
This model can be used for both regions at continental scales -- it can capture both plug and shear flow.

\item \textcolor{blue}{Section 2.2.1: in my opinion, the authors do not provide sufficient justification for the penalty term,
equation (7). They describe this as something that is added to smooth over artifacts - this would be
needed based on the discretization, which there is little discussion of. The authors should state what
order finite elements they are using - I presume they are linear, and that this is why the stabilization
is needed? Why is stabilization needed only for the SIA? I think these things should be made clear.}

See comments by reviewer \#3.
The technical answer is that this makes the solution live in the Sobolev space $H^1(\Omega)$.
A more heuristic answer is that the shallow ice approximation is usually assumed to hold only over distances greater than a few ice thicknesses.
The penalty term is meant to filter out variability at length scales where the model doesn't even apply.
\item \textcolor{blue}{Section 2.2.2: there is some imprecision here in equation (13) - you have not defined anywhere that $\Gamma$
is the boundary, and which boundary you are referring to. One can figure it out, but it is not precise.
$\Omega$ is not defined either though one will assume invariably that this is an open bounded domain in 2D
or 3D depending on which approximation one is looking at.}

@@

We also added an entirely new section which is now \S2.3 in the text just on the boundary conditions for the different diagnostic models.

\item \textcolor{blue}{Certain terms in the equations I do not believe are defined anywhere, for instance, in equation (10),
there is no expression given for the strains $\dot\epsilon(u)$.
This is one of the things I cam across that need to
be made more precise.}

This was stated in table 1 but we've added the definition as $\frac{1}{2}(\nabla u + \nabla u^\top)$ and added a sentence to the text referring readers to table 1.
\item \textcolor{blue}{Section 2.2.3: the authors comment that higher degree polynomials can be used in the vertical layer
in the hybrid approach. What order is typically used?}

Added the following text: ``Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.''
\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen's flow law come into the hybrid model?
I was expecting to see it there, but maybe I'm missing something.}

This was a bad oversight on our part -- the hybrid model does use Glen's flow law and we've added more detail at the end of this section describing the terms in the action functional (equations 22 through 26 in the revised text).
The main difference is the term for viscous power dissipation.
\item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use
different models in different regions and couple them (a la the ISCAL method). Is this possible, or
something that the authors are thinking to add to their model/code?}

### 2021-04-05 icepack.tex
Expand discussion of hybrid model, final responses

>>>>

\textbf{Action functional.}
We can now describe all of the terms in the action functional for this model:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega\int_0^1hA^{-\frac{1}{n}}\sqrt{|\dot\varepsilon(u)|^2 + h^{-2}|\partial_\zeta u|^2}^{\frac{1}{n} + 1}\,d\zeta\,dx \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u(\zeta = 0)|^{\frac{1}{m} + 1}dx \\
    \text{gravity} & = -\int_\Omega\int_0^1\rho_Igh\nabla s\cdot u\,d\zeta\,dx \\
    \text{terminus} & = \int_\Gamma\int_0^1(p_I - p_W)u\cdot\nu\;d\zeta\,d\gamma
\end{align}
where $p_I$ is the ice overburden pressure and $p_W$ is the water pressure (if any) at the calving terminus from the ocean or a pro-glacial lake.
The full action functional is exactly analogous to that of the shallow stream model:
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}
\end{equation}
Again, we note that the expression for horizontal the strain rate in terrain-following coordinates comes from equation \eqref{eq:geometric-correction}, which includes the additional geometric correction factor.

In terrain-following coordinates, the ice overburden pressure at relative depth $\zeta$ is $p_I = \rho_Igh(1 - \zeta)$.
The water pressure is zero above the waterline and increases linearly below it.
Letting $\zeta_{\text{sl}}$ denote the relative depth to the waterline, the water pressure is
\begin{equation}
    p_W = \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+
\end{equation}
where $w_+$ denotes the positive part of a real number.
This quantity is continuous but only piecewise linear.
Integrating it correctly requires more work above and beyond the usual symbolic approach; see section \S\ref{subsubsec:ocean-bc} for details.
